\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{parskip}
\usepackage{mathtools}
\usepackage{ebproof}

\DeclareUnicodeCharacter{2212}{-}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]
\newtheorem{theorem}{Thereom}[section]
\newtheorem{lemma}{Lemma}[section]

\usepackage{tikz-cd}
\usetikzlibrary{arrows}
\tikzset{
    commutative diagrams/.cd,
    arrow style=tikz,
    diagrams={>=space}
}
\tikzcdset{
    arrow style=tikz,
    diagrams={>={Straight Barb[scale=0.8]}}
}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\N}{\mathbb{N}}

\setlength{\parindent}{0pt}

\begin{document}


\section{The Most Important Definition}
\begin{definition}
    A \textit{functor} $U : \mathcal{C} \rightarrow \mathcal{D}$ has a \textit{left adjoint} if
        $$
        \forall\ X \in \mathcal{D}
        \ \exists
        \ FX \in \mathcal{C} \ \textrm{and}\ \eta_X : X \rightarrow UFX \in \mathcal{D}
        $$
        such that
        $$
        \forall\ A \in \mathcal{C}\ \textrm{and}\ \forall\ f : X \rightarrow UA
        \ \exists
        \ !g : FX \rightarrow A
        $$
        such that the following diagram commutes
    \begin{center}
        \begin{tikzcd}[sep=large]
            X \drar[swap]{f} \rar{\eta_X} & UFX \dar{Ug} \\
            & UA
        \end{tikzcd}
    \end{center}
\end{definition}
\pagebreak
\tableofcontents
\pagebreak
\section{Categories and Functors}

\begin{definition}
    A \textit{category} $\mathcal{C}$ consists of
    \begin{itemize}
        \item A set $Ob\,\mathcal{C}$, elements of which are called \textit{objects} of $\mathcal{C}$.
        \item For each $X, Y \in Ob\,\mathcal{C}$
            a set $\mathcal{C}(X,Y)$ called the \textit{homset} from $X$ to $Y$.
        \item For each $X, Y, Z \in Ob\,\mathcal{C}$ a \textit{composition function} $\circ$
            $$\mathcal{C}(X,Y) \circ \mathcal{C}(Y,Z) \rightarrow \mathcal{C}(X,Z)$$
        \item For all $X \in Ob\,\mathcal{C}$ an element $\iota_X$ of $\mathcal{C}(X,\ X)$,
            or equivalently a function $\iota_X : 1 \rightarrow \mathcal(X,\ X)$.
            Such that
            $$\forall f \in \mathcal{C}(X,Y)\ f \circ \iota_X = \iota_Y \circ f = f$$
        \item Composition is associative.
            $$\forall
              f \in \mathcal{C}(X,Y),
            \ g \in \mathcal{C}(Y,Z),
            \ h \in \mathcal{C}(X,Z)
            \ f \circ (g \circ h) = (f \circ g) \circ h
            $$
            %%\begin{center}
            %%\begin{tikzcd}
            %%X\ar[r,"f"]\ar[rr,out=-30,in=210,swap,"h"] & Y\ar[r,"g"] & Z
            %%\end{tikzcd}
            %%\end{center}
    \end{itemize}
\end{definition}

\par
An element $f$ of $\mathcal{C}(X,Y)$ is called an \textit{arrow},
or a \textit{morphism}. The object $X$ is called the \textit{domain} of $f$ and $Y$ is
the \textit{codomain}.

\begin{example}
    The foremost example of a category is the category of \textit{small sets}.
    We say small sets due to Russel's paradox, conversely phrased a category is small
    if its objects and arrow constitute sets; otherwise it is large.

    The objects in $Ob\,\mathcal{C}$ are sets,
    a morphism from $X$ to $Y$ is a function $f : X \rightarrow Y$.
    The composition of Set is given by composition of functions,
    and the identity maps are given by the identity functions.
\end{example}

\begin{definition}
    A \textit{functor} $U : \mathcal{C} \rightarrow \mathcal{D}$ consists of
    \begin{itemize}
        \item a function $Ob\,U : Ob\,\mathcal{C} \rightarrow Ob\,\mathcal{D}$.
        \item For each $X, Y \in Ob\,\mathcal{C}$ a function
            $$U_{X,Y} : \mathcal{C}(X,Y) \rightarrow \mathcal{D}(UX, UY)$$
            such that $U$ respects both composition and identity.
    \end{itemize}
\end{definition}

\begin{definition}
    A \textit{natural transformation} $\alpha : U \rightarrow V$
    given categories $\mathcal{C}$ and $\mathcal{D}$,
    with functors $U, V : \mathcal{C} \rightarrow \mathcal{D}$
    consists of
    $$
    \forall\ X \in Ob\,\mathcal{C}\ \textrm{a map}
    \ \alpha_X : UX \rightarrow VX
    $$
    such that $\forall\ f : X \rightarrow Y \in \mathcal{C}$ the following commutes
    \begin{center}
        \begin{tikzcd}[sep=large]
            UX \rar{\alpha_X} \dar[swap]{Uf} & VX \dar{Vf} \\
            UY \rar{\alpha_Y}                & VY
        \end{tikzcd}
    \end{center}
    A natural transformation can be considered a morphism of functors.
\end{definition}

\pagebreak
\section{Monoids and Groups}

\begin{definition}
    A \textit{monoid} is a set $M$ together with an associative binary operation $\cdot$ such that
    $$\forall\ x, y, z \in M\ (x \cdot y) \cdot z = x \cdot (y \cdot z)$$
    and a unit element $e \in M$ where
    $$\forall\ m \in M\ m \cdot e = e \cdot m = m$$
    Thus a monoid could be a considered a triple $(M,\ \cdot,\ e)$.
\end{definition}

At this point we can define the "forgetful functor", consider a functor from the
category of monoids to the category of sets. For this functor we can simply prise the
set $M$ from the monoid and "forget" the binary associative operation $\cdot$ as well as
the unit $e$. It is called the forgetful functor because it forgets some of the structure of
the domain category.

\begin{definition}
    Given two monoids $M$ and $N$, a \textit{monoid map} is a function $f : M \rightarrow N$
    which preserves multiplication and the unit where
    $$f(e_M) = e_N\ \textrm{and}\ \forall x, y \in M f(x \cdot y) = z, z \in N$$
\end{definition}

\begin{example}
    Perhaps the simplest example of a monoid is taking the natural numbers $\mathbb{N}$
    (including zero), addition, and zero.
    Thus the monoid is $(\mathbb{N}, +, 0)$. We can see that axioms
    are obeyed. Addition is associative and $\forall x \in \mathbb{N}\ x + 0 = x$.
    A further example is multiplication with unit element being 1.
\end{example}

\begin{definition}
    A \textit{commutative monoid} is a monoid $(M,\ \cdot,\ e)$ where
    $$\forall x, y \in M\ x \cdot y = y \cdot x$$
\end{definition}

Because both addition and multiplication are commutative, the monoids $(\mathbb{N},\ +,\ 0)$
and $(\mathbb{N},\ \times,\ 1)$ are commutative monoids.
$$ \forall x, y \in \mathbb{N}$$
$$x + y = y + x$$
$$x \times y = y \times x$$

\begin{example}
    An example of a monoid which is not commutative is that of strings and concatenation.
    Consider the alphabet $\Sigma = \{a,\ b,\ \epsilon\}$ where $\epsilon$ is the empty string.
    Strings in this alphabet include $\epsilon,\ aa,\ ab,\ bbb$ and so on.
    Our binary associative operation is concatenation ($+$).
    The unit element is $\epsilon$.
    From this we can see that
    $$aa + (bb + ab) = (aa + bb) + aa = aabbaa$$
    $$aa + \epsilon = aa = \epsilon + aa$$
    Furthermore that
    $$aa + bb \ne bb + aa$$
\end{example}

A further example of a non-commutative monoid is that of matrix multiplication.

\begin{definition}
    A \textit{idempotent monoid} is one where $\forall\ x \in M\ x \cdot x = x$.
\end{definition}

To understand this definition consider the monoid of sets and union, $(P_f(X), \cup, \varnothing)$.
$M$ is the finite power set of $X$ (set of all finite subsets), $\cdot$ is union, and $e$ is
the empty set $\varnothing$.
This is an example of an idempotent monoid because
$$\forall\ x \in P_f(X)\ x \cup x = x$$
It is also a commutative monoid because
$$\forall\ x,\ y \in P_f(X)\ x \cup y = y \cup x$$

Note that is is convention for commutative monoids to use the notation
$(M,+,0)$ rather than $(M,\cdot,e)$.

\begin{definition}
    A \textit{group} is a monoid where there exists an inverse element such that
    $$\forall\ m \in M\ \exists\ n \in M\ s.t.\ m \cdot n = e$$
\end{definition}

\begin{definition}
    An \textit{Abelian group} is a commutative group, i.e.
    $$\forall\ x, y \in M\ x \cdot y = y \cdot x$$
\end{definition}

\begin{example}
    An exmaple of an Abelian group is the integers $\mathbb{Z}$ and addition.
\end{example}

\pagebreak
\section{Left adjoints}

\begin{definition}
    An \textit{isomorphism} consists of maps
    $f : X \rightarrow Y$ and $g : Y \rightarrow X$
    such that $gf = \iota_X$ and $fg = \iota_Y$.
\end{definition}

%TODO
%Often one refers to a given map f : X −→ Y as an isomorphism. The reason one can do that is because any inverse of f is unique, i.e., if g and g′ are both inverses of f, it must be the case that g = g′.
%Often one refers to a given map f : X −→ Y as an isomorphism. The reason one can do that is because any inverse of f is unique, i.e., if g and g′ are both inverses of f, it must be the case that g = g′.
%All functors preserve isomorphisms, i.e., given any functor H : C −→ D and any isomorphism f in C, the map Hf must be an isomorphism in D.
%One needs to be careful here: people often say “X and Y are isomorphic,” and they typically think they mean that there exists an isomorphism between the objects X and Y . But in fact, they almost always have a specific isomorphism in mind. This has led to an enormity of mistakes in both papers and books in the area. For instance, any two countably infinite sets are isomorphic in the category Set, but there are uncountably many isomorphisms between any two countably infinite sets.

\begin{definition}
    A \textit{functor} $U : \mathcal{C} \rightarrow \mathcal{D}$ has a \textit{left adjoint} if
        $$
        \forall\ X \in \mathcal{D}
        \ \exists
        \ FX \in \mathcal{C} \ \textrm{and}\ \eta_X : X \rightarrow UFX \in \mathcal{D}
        $$
        such that
        $$
        \forall\ A \in \mathcal{C}\ \textrm{and}\ \forall\ f : X \rightarrow UA
        \ \exists
        \ !g : FX \rightarrow A
        $$
        such that the following diagram commutes
\end{definition}

\begin{center}
    \begin{tikzcd}[sep=large]
        X \drar[swap]{f} \rar{\eta_X} & UFX \dar{Ug} \\
        & UA
    \end{tikzcd}
\end{center}

We will now go through a proof that a functor has a left adjoint.
In this proof $\mathcal{C}$ is the category of idempotent commutative monoids,
and $\mathcal{D}$ is the category of sets.
An example of an $ICMonoid$ is that of $(P_f(X),\ \cup,\ \varnothing)$,
where $P_f(X)$ is the finite power set of $X$;
i.e. $P_f(X)$ is the set of all finite subsets of $X$.

\begin{proof}
    The forgetful functor $U : ICMonoid \rightarrow Set$,
    from the category of idempotent commutative monoids
    to the category of sets,
    has a left adjoint.

    Given an arbitrary set $X \in \mathcal{D}$,
    let $\eta_X : X \rightarrow UFX$,
    $FX$ will be $P_f(X)$

    The unit $\eta_X : X \rightarrow P_f(X)$ sends an elemenxt $x$ of $X$ to the singleton $\{x\}$.

    Consider $FX$ to be $P_f(X)$, the finite power sets



    %%TODO

\end{proof}


The forgetful functor U : ICMonoid $\rightarrow$ Set
from the category of idempotent commmutative monoids and maps of idempotent commutative monoids
to Set has a left adjoint given by putting $FX = P_f(X)$,
the set of finite subsets of X.
\vspace{5mm}

The multiplication of FX = Pf(X) is given by union,
with the unit given by the empty set.
One can readily check that these satisfy the axioms for an idempotent commutative monoid.
\vspace{5mm}

The unit $\eta_X : X \rightarrow P_f(X)$ sends an elemenxt $x$ of $X$ to the singleton $\{x\}$.
Given any idempotent commutative monoid $M$ and any function $f : X \rightarrow M$,
define $g : P_f (X) \rightarrow M$ by $g(A) = \Pi_{x \in A} f(x)$.
\vspace{5mm}

It follows from idempotence and commutativity of $M$ that $g$ is a map of monoids,
hence necessarily a map of idempotent commutative monoids.
\vspace{5mm}

It is routine to check $g\eta_X = f$.
Unicity of g is routine.

\vspace{5mm}
LINE BREAK
\vspace{5mm}

Given category of commutative monoids $\mathcal{C}$
and the category of (small) sets $\mathcal{D}$
and a functor $U : \mathcal{C} \rightarrow \mathcal{D}$
$U$ has a left adjoint.
Given an arbitrary set $X \in \mathcal{D}$,
we define $\eta_X : X \rightarrow UFX$ to be $X \rightarrow \{X\}$

\vspace{5mm}
LINE BREAK
\vspace{5mm}














\pagebreak
\section{Pre-Orders, Partially Ordered Sets, and $\omega$-cpos}

\begin{definition}
    A \textit{Pre-Order} is a set $S$ with a binary relation $\leq$ such that
    $\forall\ x,\  y,\  z \in S$
    \begin{enumerate}
        \item Reflexivity - $x \leq x$
        \item Transitivity - $x \leq y$ and $y \leq z$ means that $x \leq z$
    \end{enumerate}
\end{definition}

\begin{definition}
    A \textit{Partially Ordered Set} or a \textit{POSet} is a set $S$ and a binary relation $\leq$ such that
    $\forall\ x,\  y,\  z \in S$
    \begin{enumerate}
        \item Reflexivity - $x \leq x$
        \item Anti-symmetry - $x \leq y$ and $y \leq x$ means that $x = y$
        \item Transitivity - $x \leq y$ and $y \leq z$ means that $x \leq z$
    \end{enumerate}
\end{definition}

Immediate examples of POSets include $(\mathbb{N},\leq)$ and $(\mathbb{N},\geq)$.
This is also true for the rationals and real numbers.
We can also see that $(P(X), \subseteq)$ is a POSet.
For any set $X$ we also have simply $(X,=)$.
These are also all pre-orders.
\\ \\
An example of a pre-order which is not a POSet is that of the complex numbers
where $\leq$ compares $x = (r_1,\theta_1)$ and $y = (r_2,\theta_2)$ by $r_1 \leq r_2$;
this means that where $r_1 = r_2$ it follows that $x \leq y$ and $y \leq x$
but it does not follow that $x = y$ because it is possible that $\theta_1 \neq \theta_2$.
\\ \\
Another example if that of $(\mathbb{Z},\leq)$ where comparison of $x$ and $y$ is done by
$|x| \leq |y|$.
Comparing $2$ and $-2$ we get $2 \leq -2$ and $-2 \leq 2$ but not that $2 = -2$.

\begin{definition}
    A \textit{map of POSets} is a function $f : (P_1,\leq_1) \rightarrow (Q_1,\leq_2)$ where
    $$\forall\ x,y \in P_1\ \textrm{iff}\ x \leq_1 y\ \textrm{then}\ f(x) \leq_2 f(y)$$
\end{definition}

\begin{definition}
    A \textit{$\omega$ complete partial order} is a POSet where for every ascending chain
    $$x_1 \leq x_2 \dots \leq x_n$$
    there is a least upper bound $x_\omega$ such that $\forall\ n\ x_n \leq x_omega$.
\end{definition}

\begin{definition}
    A map of $\omega$-cpo's is called \textit{$\omega$-continuous}
    if it preserves $x_\omega$.
\end{definition}

\pagebreak
\section{Initial and Terminal Objects}

\begin{definition}
    A category $\mathcal{C}$ has a \textit{initial object} $I$ if
    $$\forall\ X \in Ob\,\mathcal{C}\ \exists\ !i : I \rightarrow X$$
\end{definition}

\begin{definition}
    A category $\mathcal{C}$ has a \textit{terminal object} $T$ if
    $$\forall\ X \in Ob\,\mathcal{C}\ \exists\ !t : X \rightarrow T$$
\end{definition}

The following all have terminal objects $1$
POSets
Pre-Orders
Monoids

\begin{theorem}
    Given categories $\C$ and $\D$,
    if $\C$ has a terminal object $T$
    and $U : \C \rightarrow \D$ has a left adjoint
    then $\D$ has a terminal object $UT$.
\end{theorem}

\begin{proof}
    Let $T$ be a terminal object of $\C$, given an arbitrary
    $X \in \D\ \exists\ !t : FX \rightarrow UT$.
    We can obtain a map $X \rightarrow UT$ through $Ut \circ \eta_X$.

    To prove this is unique take $h$ and $h\prime : X \rightarrow UT$
\end{proof}

%%Given categories C and D with terminal objects TC and TD respectively, and given a functor H : C −→ D, there is necessarily a unique map !H(TC ) from H(TC) to TD. The functor H is said to preserve the terminal object if the map
%%!H(TC) : H(TC) −→ TD
%%is an isomorphism.

\pagebreak
\section{Binary Products}
\begin{definition}
    A category $\C$ has \textit{binary products} if
    $$
    \forall\ X,\,Y \in \C
    \ \exists\ \Pi_X : X \times Y \rightarrow X
    \ \exists\ \Pi_Y : X \times Y \rightarrow Y
    $$
    such that for all diagrams of the form
    \begin{center}
        \begin{tikzcd}[sep=large]
            & A \drar{f} \dlar[swap]{g} & \\
            X & & Y
        \end{tikzcd}
    \end{center}
    there exists a unique map $h : A \rightarrow X \times Y$
    such that the following diagram commutes
    \begin{center}
        \begin{tikzcd}[sep=large]
            & A \drar{f} \dlar[swap]{g} \dar[dashed]{h} &   \\
            X & \rar{\Pi_X} X \times Y \lar{\Pi_Y}      & Y
        \end{tikzcd}
    \end{center}
\end{definition}

\begin{proof}
    The category POSet of partially ordered sets has binary products.
\end{proof}

\begin{definition}
    A category $\C$ has \textit{finite products} if
    it has binary products
    and it has a terminal object.
\end{definition}

Prove that the category CMonoid of commutative monoids has finite products.
\begin{proof}
    The category commutative monoids has a terminal object given by the one element monoid.
    The category CMonoid has binary products.
\end{proof}

\begin{definition}
    Given categories $\C$ and $\D$,
    the product of $\C$ and $\D$ is the category $\C \times \D$
    given as follows $Ob\,(\C \times \D) = Ob\,\C \times Ob\,\D$
    %%• (C×D)((X,Y),(X′,Y′))=C(X,X′)×D(Y,Y′)
    %%with composition and identities in C × D determined by those in C and D.
\end{definition}

\begin{definition}
    The \textit{diagonal functor} denoted $\Delta : \C \rightarrow \C$.
    %%Given any category C we denote by ∆ : C −→ C × C, the functor that sends an object X of C to (X,X) and does likewise to maps. This is called the diagonal functor.
\end{definition}

%%TODO Theorem 5.10 If a functor U : C −→ D has a left adjoint, U preserves any finite products that exist in C.

\pagebreak
\section{Coproducts}

\pagebreak
\section{Right Adjoints}
\begin{definition}
    A \textit{functor} $U : C \rightarrow D$ has a \textit{right adjoint} if
        $$
        \forall\ X \in D
        \ \exists
        \ RX \in \mathcal{C}\ \textrm{and}\ \epsilon_X : URX \rightarrow X \in D
        $$
        such that
        $$
        \forall\ A \in \mathcal{C}\ \textrm{and}\ \forall\ f : UA \rightarrow X
        \ \exists
        \ !k : A \rightarrow RX
        $$
        such that the following diagram commutes
    \begin{center}
        \begin{tikzcd}[sep=large]
            X & URX \lar[swap]{\epsilon_X}\\
              & UA \ular{h} \uar[swap]{k}
        \end{tikzcd}
    \end{center}
\end{definition}
\subsection{Exponential Object}

\pagebreak
\section{Cartesian closed category}

\begin{definition}
    A category $\C$ is called \textit{cartesian closed}
    if it has finite products
    and the functor %%TODO
    blank cross x c to c has a right adjoint.
\end{definition}

\subsection{Proof Set is Cartesian Closed}
\begin{proof}
    First observe that \textit{Set}, the category of sets,
    has a terminal object given by any one element set.
    Binary products of the sets $X$ and $Y$ are given by the
    cartesian product $X \times Y$,
    together with the projections to $X$ and $Y$.
    Given a set $X$, the functor $- \times X : Set \rightarrow Set$,
    has a right adjoint given by
    sending a set $Y$ to the set $X \Rightarrow Y$ of all functions from $X$ to $Y$,
    with counit
    $$\epsilon_X : ((X \Rightarrow Y) \times X) \rightarrow Y$$
    given by evaluation i.e. $\epsilon_X(h,x) = h(x)$.
\end{proof}

\subsection{Proof PreOrd is Cartesian Closed}
\begin{proof}
    First observe that a terminal object given by the trivial one element pre-order.
    Then observe it has binary products given by cartesian product $P \times Q$ with pointwise order,
    together with the projections to $P$ and $Q$.
    Given a pre-order $(P,\leq_P)$ we need to prove the functor
    $$- \times (P,\leq_P) : PreOrd \rightarrow PreOrd$$
    has a right adjoint.
    Given a pre-order $(Q,\leq_Q)$, define $P \Rightarrow Q$ to be
    the set of ordering preserving functions from $P$ to $Q$ and put
    $$h \leq_{P \Rightarrow Q} h\prime$$
    if $\forall x \in P$ one has $h(x) \leq_Q h\prime(x)$.
    Observe that $(P \Rightarrow Q, \leq_{P \Rightarrow Q})$ is a pre-order.
    Define $\epsilon_P : (P \Rightarrow Q) \times P \rightarrow Q$ by evaluation;
    i.e. $\epsilon_P(h, x) = h(x)$,
    observe that $\epsilon_P$ is order preserving.
    Given any order preserving function of the form
    $$f : R \times P \rightarrow Q$$
    it follows from the cartesian closedness of Set that there is a unique
    function $g$ from $R$ to the set of all functions from $P$ to $Q$
    which makes the requisite diagram commute,
    that function sending $z \in R$ to $f(z,−) : P \rightarrow Q$.
    Finally observe that,
    firstly for each $z$ the function $f(z,-)$ is order preserving,
    and secondly if $z \leq_R z\prime$ then $f(z,-) \leq_{P \Rightarrow Q} f(z\prime, -)$.


\end{proof}

\subsection{Proof POSet is Cartesian Closed}
\begin{proof}
    First observe that \textit{Set}, the category of sets,
    has a terminal object given by any one element set.
    Binary products of the sets $X$ and $Y$ are given by the
    cartesian product $X \times Y$,
    together with the projections to $X$ and $Y$.
    Given a set $X$, the functor $- \times X : Set \rightarrow Set$,
    has a right adjoint given by
    sending a set $Y$ to the set $X \Rightarrow Y$ of all functions from $X$ to $Y$,
    with counit
    $$\epsilon_X : ((X \Rightarrow Y) \times X) \rightarrow Y$$
    given by evaluation i.e. $\epsilon_X(h,x) = h(x)$.
\end{proof}

\subsection{Proof $\omega$-CPOs is Cartesian Closed}

\begin{lemma}
    Given a doubly-indexed sequence
    $(x_{nm})n\in N,m\in N$
    of elements of an $\omega$-cpo such that for all $n$ and $m$ we have,
    $$x_{n0} \leq x_{n1} \leq x_{n2} \leq \dots\quad\textrm{and}\quad
    x_{m0} \leq x_{m1} \leq x_{m2} \leq \dots$$
    then
    $$lub_mlub_n(x_{nm}) = lub_nlub_m(x_{nm}) = lub_k(x_{kk})$$
\end{lemma}

%Given ω-cpo’s P and Q,
%let ωCpo(P,Q) denote the set of maps of ω-cpo’s from P to Q.
%Consider the partial order structure on ωCpo(P,Q)
%givenbyputtingf≤gifforallxinP,onehasf(x)≤g(x)inQ.
%Using Lemma , prove that this partial order has
%least upper bounds of ω-chains and that evaluation respects them.

\begin{proof}

\end{proof}

\subsection{Proof Monoid is not Cartesian Closed}
\begin{proof}
    Consider that if a functor has a right adjoint it must preserve finite coproducts,
    hence the initial object.
    The initial object of the category of monoids Monoid is given by the one element monoid 1.
    Thus if Monoid was cartesian closed it would follow that for every monoid M,
    1 would be isomorphic to $1 \times M$;
    the underlying set would be the cartesian product of 1 and $M$.
    Let $M$ be the natural numbers,
    the cartesian product of $1 \times M$ would have the cardinality of the natural numbers
    thus is it not possible for it to be isomorphic to 1.
    Hence Monoid is not cartesian closed.
\end{proof}
An alternate proof is
\begin{proof}
    Monoid is not cartesian closed because there exists a monoid $M$ such that
    the functor $- \times M$ does not have a right adjoint.
    It does not have a right adjoint because right adjoints preserve finite coproducts.
    To preserve finite coproducts a functor must preserve the initial object and coproducts.
    The initial object of Monoid is the one element monoid 1.
    Consider then that $1 \times M$ is not isomorphic to the initial object 1,
    when we have $M$ be any non-trivial monoid, for example $\N$.
    Thus because $M \times 1$ is not isomorphic to the initial object
    the functor $- \times M$ does not preserve initial objects,
    thus not finite coproducts,
    thus it does not have a right adjoint
    thus Monoid is not cartesian closed.
    $$1 \times M \ncong 1$$
\end{proof}
We can also consider that in a cartesian closed category,
for example Set where the initial object is the empty set $\varnothing$,
we have $\varnothing \times X = \varnothing$ thus preserving the initial object.

\pagebreak
\section{Simply Typed Lambda Calculus}
\begin{definition}
    The \textit{simply-typed $\lambda$-calculus} consists of
    types and terms;
    defined as such
    \begin{itemize}
        \item Types
            $T ::= B 
            \ | \ 1
            \ | \ T_1 \times T_2
            \ | \ T \Rightarrow T \prime$
        \item Terms
            $t ::=  x
            \ |\ \star
            \ |\ f(t_1,\dots t_n)
            \ |\ \langle t_1, t_2 \rangle
            \ |\ \pi_i(t)
            \ |\ \lambda x.t
            \ |\ t_1 t_2$
    \end{itemize}
\end{definition}

\vspace{4mm}
Next we must define a context to equate terms to types.

\begin{definition}
A \textit{context} $\Gamma$ consists of a finite list of distinct variables each associated with types,
    typically read as saying $x_i$ is of type $T_i$ for each $i$.
    $$\Gamma =x_1 :T_1,\dots,x_n :T_n$$
\end{definition}

\subsection{Well-Formedness Judgements}
Well-formedness judgements consist of a context $\Gamma$ together with a term,
typically written as $x_1 :T_1,\dots, x_n :T_n \vdash t : T$ or $\Gamma \vdash t:T$
generated by the following rules 

Variable
$$
\frac{}{t_1 : T_1, \dots t_n : T_n \vdash t_i : T_i}
$$

Unit
$$
\frac{}{\Gamma \vdash \star : 1}
$$

Function
$$
\frac{\Gamma \vdash t_1 : T_1 \dots t_n : T_n}{\Gamma \vdash f(t_1,\dots t_n) : T}
\quad\textrm{where}\ f : T_1 \times T_2 \times \dots T_n \rightarrow T
$$

Binary Product
$$
\frac
{\Gamma \vdash t_2 : T_1 \quad \Gamma \vdash t_2 : T_2}
{\Gamma \vdash \langle t_1, t_2 \rangle : T_1 \times T_2}
\quad
\frac
{\Gamma \vdash t : T_1 \times T_2}
{\Gamma \vdash \pi_1(t) : T_1 \quad \Gamma \vdash \pi_2(t) : T_2}
$$

$\lambda$-Abstraction
$$
\frac
{\Gamma, x : T \vdash t : T\prime}
{\Gamma, x : T \vdash \lambda x.t : T \Rightarrow T\prime}
$$

Application
$$
\frac
{\Gamma \vdash t_1 : T \Rightarrow T\prime \quad \Gamma \vdash t_2 : T}
{\Gamma \vdash t_1 t_2 : T\prime}
$$

\subsection{Equality}
The equality judgements of central importance are

$\alpha$-equality
$$
\frac
{\Gamma, x : T \vdash t : T\prime}
{\Gamma \vdash \lambda x.t = \lambda x\prime.t[x/x\prime] : T \Rightarrow T\prime}
\quad
\textrm{if $x\prime$ does not appear in $t$}
$$

$\beta$-equality
$$
\frac
{\Gamma, x : T \vdash t\prime : T\prime \quad \Gamma \vdash t : T}
{\Gamma \vdash (\lambda x.t\prime)t = t\prime[t/x]: T\prime }
$$

$\eta$-equality
$$
\frac
{\Gamma \vdash t : T \Rightarrow T\prime}
{\Gamma \vdash \lambda x.(tx) = t : T \Rightarrow T\prime}
\quad
\textrm{if $x$ does not appear in $t$}
$$

\subsection{Rules}
From our well-formedness judgements we can derive three rules
\vspace{4mm}
Exchange
$$
\frac
{\Gamma \vdash t : T}
{\pi\Gamma \vdash t : T}
\quad
\textrm{where }
\pi \Gamma
\textrm{ is any permutation of}
\Gamma
$$
Weakening
$$
\frac
{\Gamma \vdash t : T}
{\Gamma, x : T\prime \vdash t : T}
$$
Contraction
$$
\frac
{\Gamma, y : T, z : T \vdash t : T\prime}
{\Gamma, x : T \vdash t[x/y][x/z] : T\prime}
$$

\begin{proof}
    The \textit{weakening rule} is derived as such
    Proof of (Variable) and (Unit) are trivial.
    (Function)
    \begin{align*}
         \textrm{Given a proof of }\Gamma \vdash f(t_1,\dots t_n) : T\ \exists \textrm{ proof of }
         &\Gamma \vdash t_1 : T_1, \dots \Gamma \vdash t_n : T_n \\
         \textrm{where }
         &\Gamma \vdash f : T_1 \times \dots T_n \rightarrow T \\
         \textrm{By induction } \exists \textrm{ proof of }
         &\Gamma, x : T\prime \vdash t_1 : T_1, \dots \Gamma, x : T\prime \vdash t_n : T_n\\
         \textrm{Append (function) to obtain proof of }
         &\Gamma, x : T\prime \vdash f(t_1, \dots t_n) : T\\
    \end{align*}

    (Binary Product)
    \begin{align*}
        \textrm{The same as (Function)}
    \end{align*}

    ($\lambda$ Abstraction)
    \begin{align*}
         \textrm{Given a proof of }\Gamma \vdash \lambda x.t : T \Rightarrow T\prime\ \exists \textrm{ proof of }
        &\Gamma, x : T \vdash t : T\prime \\
         \textrm{By induction } \exists \textrm{ proof of }
        &\Gamma, x : T, y : T\prime\prime \vdash t : T\prime \\
         \textrm{Using the }\mathit{exchange\ rule\ } \exists \textrm{ proof of }
        &\Gamma, y : T\prime\prime, x : T \vdash  t : T\prime \\
         \textrm{Then append }(\lambda)\textrm{ for proof of }
        &\Gamma, y : T\prime\prime \vdash \lambda x.t : T \Rightarrow T\prime
    \end{align*}
    The exchange rule can be derived from the well-formedness judgements.
    It is defined as such
    $$
    \frac
    {\Gamma \vdash t : T}
    {\pi\Gamma \vdash t : T}
    \quad
    \textrm{where }
    \pi \Gamma
    \textrm{ is any permutation of}
    \Gamma
    $$

    (Application)
    \begin{align*}
        \textrm{Given a proof of } \Gamma \vdash t_1 t_2 : T\ \exists \textrm{ a proof of }
        &\Gamma \vdash t_1 : T \Rightarrow T\prime \textrm{ and } \Gamma \vdash t_2 : T\\
        \textrm{By induction we have a proof of }
        &\Gamma, x : T\prime\prime \vdash t_1 : T \Rightarrow T\prime \\
        \textrm{ and }
        &\Gamma, x : T\prime\prime \vdash t_2 : T \\
        \textrm{Append (Application) to obtain proof of }
        &\Gamma, x : T\prime\prime \vdash t_1 t_2 : T\prime
    \end{align*}
\end{proof}

\subsection{Models}
\begin{definition}
    Given a cartesian closed category $\C$,
    base types $B$,
    function symbols $f$,
    equations between them,
    and assignments $M$ for $B$ and $f$ that respect both typing and equations,
    the induced model for the simply typed $\lambda$-calculus is defined inductively on types by
    \begin{itemize}
        \item $M(B)$ is given.
        \item $M(1) =$ 1 the terminal object of $\C$
        \item $M(T_1 \times T_2)$ is $M(T_1) \times M(T_2)$;
            defined using the binary product of $\C$.
        \item $M(T \Rightarrow T\prime)$ is $M(T) \Rightarrow M(T\prime)$;
            defined using the closed structure of $\C$.
    \end{itemize}
    Variable
    $$
    \frac{}{t_1 : T_1, \dots t_n : T_n \vdash t_i : T_i}
    $$
    Unit
    $$
    \frac{}{\Gamma \vdash \star : 1}
    $$
    Function
    $$
    \frac{\Gamma \vdash t_1 : T_1 \dots t_n : T_n}{\Gamma \vdash f(t_1,\dots t_n) : T}
    \quad\textrm{where}\ f : T_1 \times T_2 \times \dots T_n \rightarrow T
    $$

    Binary Product
    $$
    \frac
    {\Gamma \vdash t_2 : T_1 \quad \Gamma \vdash t_2 : T_2}
    {\Gamma \vdash \langle t_1, t_2 \rangle : T_1 \times T_2}
    \quad
    \frac
    {\Gamma \vdash t : T_1 \times T_2}
    {\Gamma \vdash \pi_1(t) : T_1 \quad \Gamma \vdash \pi_2(t) : T_2}
    $$

    $\lambda$-Abstraction
    $$
    \frac
    {\Gamma, x : T \vdash t : T\prime}
    {\Gamma, x : T \vdash \lambda x.t : T \Rightarrow T\prime}
    $$

    Application
    $$
    \frac
    {\Gamma \vdash t_1 : T \Rightarrow T\prime \quad \Gamma \vdash t_2 : T}
    {\Gamma \vdash t_1 t_2 : T\prime}
    $$
\end{definition}

\section{TODO}
\subsection{Exponential object}
\subsection{Semi-groups}
\subsection{natural transformations}
\subsection{equalisers and coequalisers}
\subsection{pullbacks and pushouts}
\pagebreak

We can derive a tree for the term $\lambda x. \lambda y.(xy)(xy)$
showing that there exists a contradiction thus it has no type.

\begin{prooftree}
    \hypo{ \Gamma &\vdash xy : R \Rightarrow T\prime\prime }
    \hypo{ \Gamma &\vdash xy : T\prime\prime }
    \infer2[(App)]{ \Gamma &\vdash (xy)(xy) : T\prime\prime }
    \infer0[(Var)]{ \Gamma &\vdash y : T\prime }
    \infer2[($\lambda$)]{ \Gamma &\vdash \lambda y.(xy)(xy) : T\prime \Rightarrow T\prime\prime }
    \infer0[(Var)]{ \Gamma &\vdash x : T }
    \infer2[($\lambda$)]{ \Gamma \vdash \lambda x. \lambda y.(xy)(xy) : T \Rightarrow T\prime \Rightarrow T\prime\prime }
\end{prooftree}

\end{document}
